\texttt{davos}: a Python package ``smuggler'' for constructing lightweight reproducible notebooks

### 2022-11-14 paper/main.tex
illustrative example walkthrough up to line 24

After importing \texttt{davos} (line 1), the user first smuggles two utilities for interacting with local files in the code below. The \texttt{smuggle} statement in line 3 loads the \texttt{is\_file()} function from the Python standard library's \texttt{os.path} module. Standard library modules are included with all Python distributions, so this line is functionally equivalent to an \texttt{import} statement and no onion comment is needed. Line 4 loads the \texttt{joblib} library~\cite{Varo10}, installing it first, if necessary. Since \texttt{joblib}'s I/O interface has historically remained stable and backwards-compatible across releases, requiring that users have a particular exact version installed would likely be unnecessarily restrictive. However, there's also no guarantee that a \textit{future} release won't introduce some breaking change, so to help ensure the analysis notebook continues to run properly in the future, the onion comment in line 4 limits allowable versions to those already released when the code was written.
%After importing \texttt{davos} (line 1), the user first smuggles two utilities that will be used to help find and load local files in the code below: the \texttt{is\_file} function from the \texttt{os.path} module, and the \texttt{joblib} library~\cite{Varo10} (lines 3--4). The first \texttt{smuggle} statement needs no onion comment, as the \texttt{os.path} module is part of Python's standard library and is therefore guaranteed to be available. The second ensures \texttt{joblib} is installed before loading it Enforcing an exact version of \texttt{joblib} is likely unnecessary, since the package's I/O interface has historically remained generally stable across versions. However, there's no

Line 6 then uses the \texttt{davos.config} object to enable \texttt{davos}'s \texttt{auto\_rerun} option before smuggling the next two packages: \texttt{NumPy}~\cite{HarrEtal20} and \texttt{pandas}. Because these libraries rely heavily on custom C data types, loading the particular versions from the onion comments may require restarting the notebook kernel if different versions were already imported during the same interpreter session (see Section~\ref{subsec:config}). This is unlikely to happen with \texttt{NumPy}, as it is loaded explicitly only once (by line 7) and is not required by any packages loaded before it (and therefore not implicitly imported while loading them). However, it's possible a future user could edit this notebook to add additional code before line 7, run the notebook's cells out of order, or even configure their IPython kernel to execute custom code on startup (e.g., as Google Colab notebooks do), any of which could potentially import a different, existing version of \texttt{NumPy} before \texttt{davos} smuggles the version specified in the onion comment. Enabling the \texttt{auto\_rerun} option before smuggling \texttt{NumPy} helps to simplify running the notebook in these scenarios, but the main benefit it affords comes when smuggling the \texttt{pandas} library.

The dataset that the user wants to analyze was provided as a \texttt{pandas.Panel} object, serialized with Python's built-in \texttt{pickle} module. The \texttt{pickle} protocol is a popular choice for persisting data in Python, allowing users to save, share, and load arbitrary objects, though with an important caveat: 
In order to successfully ``unpickle'' (i.e., load and restore) a ``pickled'' object, its class must be defined in and importable from the same module as when it was saved.
%A ``pickled'' object's class definition must be importable from the same module as when it was saved in order to successfully ``unpickle'' (i.e., load and restore) it later. 
The \texttt{Panel} class was removed from \texttt{pandas} in July, 2019 (v0.25.0), meaning that the dataset can be loaded only if the user's notebook uses a \texttt{pandas} version released prior to that change. However, more recent updates to the \texttt{pandas} library since then have brought substantial improvements that the user would like to take advantage of in their analysis code, including faster performance, more memory-efficient data types, new functions and methods, compatibility with newer versions of Python and popular plotting libraries, and better display formats for data structures (including in Jupyter notebooks).
%Recent updates to the \texttt{pandas} library have brought substantial improvements that the user would like to take advantage of in their analysis code, including faster performance, more memory-efficient data types, new functions and methods, compatibility with newer versions of Python and popular plotting libraries, and better display formats (including in Jupyter notebooks). However, the \texttt{Panel} class was removed from \texttt{pandas} in July, 2019 (v0.25.0), meaning that the dataset they are working with can only be loaded by a version released prior to that change.
\texttt{davos} makes it possible to do both within a single notebook, and with the \texttt{auto\_rerun} option, this process can be fully automated. The first time the notebook is run, lines 10--12 will install and load an older version of \texttt{pandas} that defines the \texttt{Panel} class, use it to read the dataset from its serialized file format, convert the dataset from a \texttt{Panel} to a multi-index \texttt{DataFrame}, and write its contents to a CSV file. The \texttt{smuggle} statement on line 14 will then install a newer \texttt{pandas} version (v1.3.5) for use in the analyses themselves. Because an older \texttt{pandas} version was previously loaded (by line 10), and because the package's C extension modules were modified between the old and new versions, switching to the new version will require restarting the notebook kernel. When \texttt{davos} determines this, it will automatically restart the kernel, re-run the notebook up through the current cell, and also re-queue any subsequent cells that were queued before the restart was triggered. This second time the code is run (as well as any future times), lines 10--12 will be skipped since the CSV-formatted dataset has already been created, and the \texttt{smuggle} statements up through line 14 will execute (virtually) as fast as regular \texttt{import} statements since the required package versions have already been installed.
%This second time the code is run (as well as any future times), the \texttt{smuggle} statements that have already been run once will execute (virtually) as fast as \texttt{import} statements since the required package versions are now installed, and lines 10--12 will be skipped altogether since the CSV-formatted dataset has already been created. It's worth noting that one could instead save the converted \texttt{DataFrame} as a pickle file

Line 16 then uses the \texttt{davos.configure()} function to disable the \texttt{auto\_rerun} option and simultaneously enable two other options: \texttt{suppress\_stdout} and \texttt{noninteractive}. With these options enabled, lines 17--18 \texttt{smuggle} \texttt{TensorFlow}~\cite{AbadEtal15}, a powerful end-to-end platform for building and working with machine learning models, and \texttt{UMAP}~\cite{McInEtal18b}, a library that implements a family of related manifold learning techniques. The onion comment in line 18 also specifies that \texttt{UMAP} should be installed with the optional requirements needed for its ``plot'' and ``parametric\_umap'' features. Together, these two packages depend on 36 other unique packages, most of which have dependencies of their own. And if many of these are not already installed in the user's environment, lines 17--18 could take multiple minutes to run. Enabling the \texttt{noninteractive} option ensures the installation process doesn't require any input from the user to finish successfully, allowing them to simply ignore the notebook during this time and either work in another browser tab or application, or step away from their computer entirely. Additionally, installing more than three dozen packages could result in the installer program writing an extremely large amount of standard output text to the notebook cell's output area. Enabling \texttt{suppress\_stdout} hides the output from these two \texttt{smuggle} statements so that other potentially important output (e.g., from previous \texttt{smuggle} statements) isn't buried and lost. Importantly, if either line fails for any reason, both the stdout and stderr streams will be shown alongside the regular Python traceback.

%And if many of these are not already installed in the user's environment, lines 17--18 could take multiple minutes to run and result in the installer program writing an extremely large amount of standard output text to the notebook cell's output area. Enabling the \texttt{noninteractive} option ensures the installation process doesn't require any input from the user to finish successfully, allowing them to simply ignore the notebook during this time and either work in another browser tab or application, or step away from their computer entirely. Enabling the \texttt{suppress\_stdout} option hides the output from these two \texttt{smuggle} statements so that other potentially important output (e.g., from previous \texttt{smuggle} statements) isn't buried and lost. Importantly, if either line fails for any reason, both the stdout and stderr streams will be shown alongside the regular Python traceback.

%a large amount of standard output text generated by the installer program. The \texttt{noninteractive} option ensures the installation process doesn't require any input from the user to finish successfully, allowing them to simply ignore the notebook during this time and either work in another browser tab or application, or step away from their computer entirely.

%And if many of these are not already installed in the user's environment, lines 17--18 could take multiple minutes to run. The \texttt{noninteractive} option ensures the installation process doesn't require any input from the user to finish successfully, allowing them to simply ignore the notebook during this time and either work in another browser tab or application, or step away from their computer entirely. Additionally, in the process

%And if many of these are not already installed in the user's environment, lines 17--18 could take multiple minutes to run, and generate a large amount of standard output text while doing so. The \texttt{suppress\_stdout} option hides the output from these two \texttt{smuggle} statements so that other potentially important output (e.g., from previous \texttt{smuggle} statements) isn't buried and lost. Importantly, if either line fails for any reason, both the stdout and stderr streams from the installer program will be shown alongside the regular Python traceback.

% The \texttt{suppress\_stdout} option hides the output from the installer program for these two \texttt{smuggle} statements so that other potentially important output (e.g., from previous \texttt{smuggle} statements) isn't buried and lost. Importantly, if either line fails for any reason, both its stdout and stderr streams will be shown along with the

% have 36 unique dependencies, most of which require other packages, in turn.

% and \texttt{UMAP}~\cite{McInEtal18b}. \texttt{TensorFlow} provides an end-to-end platform for building and working with machine learning models, and \texttt{UMAP} is a library for non-linear dimensionality reduction. the onion comment in line 18 also specifies that

%These options are enabled while smuggling the following two libraries \texttt{TensorFlow}~\cite{AbadEtal15}, a powerful, heavyweight machine learning

%The next two packages smuggled, \texttt{TensorFlow}~\cite{AbadEtal15} and \texttt{UMAP}~\cite{McInEtal18b} (lines 17--18), are machine learning libraries that depend on a relatively large number of heavyweight packages. The onion comment on line 18 also specifies that \texttt{UMAP} should be installed with

% The following two lines then \texttt{smuggle} \texttt{TensorFlow}~\cite{AbadEtal15}, a large library

Next, the user installs a pair of packages that extend the Jupyter notebook interface with various interactive JavaScript widgets.

>>>>

After importing \texttt{davos} (line 1), the user first smuggles two utilities for interacting with local files in the code below. The \texttt{smuggle} statement in line 3 loads the \texttt{is\_file()} function from the Python standard library's \texttt{os.path} module. Standard library modules are included with all Python distributions, so this line is functionally equivalent to an \texttt{import} statement and does not need or benefit from an onion comment. Line 4 loads the \texttt{joblib} library~\cite{Varo10}, installing it first, if necessary. Since \texttt{joblib}'s I/O interface has historically remained stable and backwards-compatible across releases, requiring that users have a particular exact version installed would likely be unnecessarily restrictive. However, there's also no guarantee that a \textit{future} release won't introduce some breaking change, so to help ensure the analysis notebook continues to run properly in the future, the onion comment in line 4 limits allowable versions to those already released when the code was written.
%After importing \texttt{davos} (line 1), the user first smuggles two utilities that will be used to help find and load local files in the code below: the \texttt{is\_file} function from the \texttt{os.path} module, and the \texttt{joblib} library~\cite{Varo10} (lines 3--4). The first \texttt{smuggle} statement needs no onion comment, as the \texttt{os.path} module is part of Python's standard library and is therefore guaranteed to be available. The second ensures \texttt{joblib} is installed before loading it Enforcing an exact version of \texttt{joblib} is likely unnecessary, since the package's I/O interface has historically remained generally stable across versions. However, there's no

Line 6 then uses the \texttt{davos.config} object to enable \texttt{davos}'s \texttt{auto\_rerun} option before smuggling the next two packages: \texttt{NumPy}~\cite{HarrEtal20} and \texttt{pandas}. Because these libraries rely heavily on custom C data types, loading the particular versions from the onion comments may require restarting the notebook kernel if different versions were already imported during the same interpreter session (see Section~\ref{subsec:config}). This is unlikely to happen with \texttt{NumPy}, as it is loaded explicitly only once (by line 7) and is not required by any packages loaded before it (and therefore not imported and cached while loading them). However, it's possible a future user could edit this notebook to add additional code before line 7, run the notebook's cells out of order, or even configure their IPython kernel to execute custom code on startup (e.g., as Google Colab notebooks do), any of which could potentially import a different, existing version of \texttt{NumPy} before \texttt{davos} smuggles the version specified in the onion comment. Enabling the \texttt{auto\_rerun} option before smuggling \texttt{NumPy} will help to simplify running the notebook in these scenarios, but the main benefit it affords comes when smuggling the \texttt{pandas} library.

The dataset that the user wants to analyze was provided as a \texttt{pandas.Panel} object, serialized with Python's built-in \texttt{pickle} module. The \texttt{pickle} protocol is a popular choice for persisting data in Python, allowing users to save, share, and load arbitrary objects, though with an important caveat:
In order to successfully ``unpickle'' (i.e., load and restore) a ``pickled'' object, its class must be defined in and importable from the same module as when it was saved.
%A ``pickled'' object's class definition must be importable from the same module as when it was saved in order to successfully ``unpickle'' (i.e., load and restore) it later.
The \texttt{Panel} class was removed from \texttt{pandas} in July, 2019 (v0.25.0), meaning that the dataset can be loaded only if the user's notebook uses a \texttt{pandas} version released prior to that change. However, more recent updates to the \texttt{pandas} library since then have brought substantial improvements that the user would like to take advantage of in their analysis code, including faster performance, more memory-efficient data types, new functions and methods, compatibility with newer versions of Python and popular plotting libraries, and better display formats for data structures (including in Jupyter notebooks).
%Recent updates to the \texttt{pandas} library have brought substantial improvements that the user would like to take advantage of in their analysis code, including faster performance, more memory-efficient data types, new functions and methods, compatibility with newer versions of Python and popular plotting libraries, and better display formats (including in Jupyter notebooks). However, the \texttt{Panel} class was removed from \texttt{pandas} in July, 2019 (v0.25.0), meaning that the dataset they are working with can only be loaded by a version released prior to that change.
\texttt{davos} makes it possible to do both within a single notebook, and with the \texttt{auto\_rerun} option, this process can be fully automated. The first time the notebook is run, lines 10--12 will install and load an older version of \texttt{pandas} that defines the \texttt{Panel} class, use it to read the dataset from its serialized file format, convert the dataset from a \texttt{Panel} to a multi-index \texttt{DataFrame}, and write its contents to a CSV file. The \texttt{smuggle} statement on line 14 will then install a newer \texttt{pandas} version (v1.3.5) for use in the analyses themselves. Because an older \texttt{pandas} version was previously loaded (by line 10), and because the package's C extension modules were modified between the old and new versions, switching to the new version will require restarting the notebook kernel. When \texttt{davos} determines this, it will automatically restart the kernel, re-run the notebook up through the current cell, and also re-queue any subsequent cells that were queued before the restart was triggered. This second time the code is run (as well as in any future runs), lines 10--12 will be skipped since the CSV-formatted dataset has already been created, and the \texttt{smuggle} statements up through line 14 will execute (virtually) as fast as regular \texttt{import} statements since the required package versions have already been installed.
%This second time the code is run (as well as any future times), the \texttt{smuggle} statements that have already been run once will execute (virtually) as fast as \texttt{import} statements since the required package versions are now installed, and lines 10--12 will be skipped altogether since the CSV-formatted dataset has already been created. It's worth noting that one could instead save the converted \texttt{DataFrame} as a pickle file

Line 16 then uses the \texttt{davos.configure()} function to disable the \texttt{auto\_rerun} option and simultaneously enable two other options: \texttt{suppress\_stdout} and \texttt{noninteractive}. With these options enabled, lines 17--18 \texttt{smuggle} \texttt{TensorFlow}~\cite{AbadEtal15}, a powerful end-to-end platform for building and working with machine learning models, and \texttt{UMAP}~\cite{McInEtal18b}, a library that implements a family of related manifold learning techniques. The onion comment in line 18 also specifies that \texttt{UMAP} should be installed with the optional requirements needed for its ``plot'' and ``parametric\_umap'' features. Together, these two packages depend on 36 other unique packages, most of which have dependencies of their own. And if many of these are not already installed in the user's environment, lines 17--18 could take multiple minutes to run. Enabling the \texttt{noninteractive} option ensures the installation process won't require any input from the user to finish successfully, allowing them to simply ignore the notebook during this time and either work in another browser tab or application, or step away from their computer entirely. Additionally, installing more than three dozen packages could result in the installer program writing an extremely large amount of standard output text to the notebook cell's output area. Enabling \texttt{suppress\_stdout} hides the output from these two \texttt{smuggle} statements so that other potentially important output (e.g., from previous \texttt{smuggle} statements) isn't buried and lost. Importantly, if either line fails for any reason, both the stdout and stderr streams will be shown alongside the regular Python traceback.
%And if many of these are not already installed in the user's environment, lines 17--18 could take multiple minutes to run and result in the installer program writing an extremely large amount of standard output text to the notebook cell's output area. Enabling the \texttt{noninteractive} option ensures the installation process doesn't require any input from the user to finish successfully, allowing them to simply ignore the notebook during this time and either work in another browser tab or application, or step away from their computer entirely. Enabling the \texttt{suppress\_stdout} option hides the output from these two \texttt{smuggle} statements so that other potentially important output (e.g., from previous \texttt{smuggle} statements) isn't buried and lost. Importantly, if either line fails for any reason, both the stdout and stderr streams will be shown alongside the regular Python traceback.

After re-enabling these two options (line 19), the user next smuggles specific versions of three plotting libraries: \texttt{Matplotlib}~\cite{Hunt07}, \texttt{seaborn}~\cite{Wask21}, and \texttt{Quail}~\cite{HeusEtal17} (lines 21--23). Because the first two are requirements of \texttt{UMAP}'s optional ``plot'' feature, they will have already been installed by line 18, though possibly as different versions than those specified in the onion comments on lines 21 and 22. If the installed and specified versions are the same, these \texttt{smuggle} statements will function like regular \texttt{import} statements to load the libraries into the notebook namespace. If they differ, \texttt{davos} will download the requested versions in place of the installed versions before doing so. It's also worth noting that while loading \texttt{UMAP} in line 18 will not result in either \texttt{Matplotlib} or \texttt{seaborn} being imported internally due to the specific layout of the \texttt{UMAP} package, if it had, \texttt{davos} would also gracefully replace the cached versions of the two packages so that the smuggled versions are loaded properly. And (as with all other \texttt{smuggle} statements in Figure~\ref{fig:illustrative-example}), if for any reason the specified package versions are incompatible with any other packages (e.g., \texttt{UMAP}), the installer program would display a warning to this effect in the notebook.

The third plotting library smuggled, \texttt{Quail}, is also pinned to a precise version with an onion comment, though in a slightly different manner than other packages in Figure~\ref{fig:illustrative-example}. Line 23 installs and loads \texttt{Quail} from the user's fork (``myfork'') of the package's GitHub repository, at a specific point in its revision history (i.e., the commit whose short hash is ``6c847a4''). This can serve as a useful way to access a version of a package the user has customized, patched, or somehow modified for this specific use case, while still allowing them to further modify their fork of the package in the future without affecting the version used in these analyses.
